Empirical Evaluation of eval in R

### 2020-08-14 inst/analysis/analysis_stable.Rmd
Add eval count and callsite distribution analysis

library(tidyverse)
library(stringr)
library(viridis)
library(vctrs)
library(DT)
library(fst)

@@

@@

@@

For the number of call sites, we either use the `srcref` when there is one. When there is not (such as in *base*), we use a combination of `caller_package` and `caller_function`.

```{r}
nb_call_sites <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)  + eval_calls %>% filter(is.na(eval_call_srcref)) %>% select(caller_package, caller_function)%>% n_distinct()
```

<details>
  <summary>More information about when there is no <code>srcref</code> </summary>

  We assume that:

  - the combination `caller_package` and `caller_function` is unique
  - there is only one call site per such `caller_function`

  It means that we get a lower bound.
</details>

There are `r nb_call_sites` different `eval` call sites.

### Per package

@@

## Number of calls to `eval`

```{r}
nb_eval_calls <- eval_calls %>% nrow()
```

There are `r nb_eval_calls` eval calls.

>>>>

library(tidyverse)
library(stringr)
library(viridis)
library(vctrs)
library(DT)
library(fst)
library(fs)

@@

The analysis is being performed on dataset read from file `r params$data_filepath`.

```{r file-info, echo=FALSE}
file_info(params$data_filepath)
```

@@

# Corpus

TODO

@@

There are four kinds of `eval` in the `base` package. Their definitions are shown below.

```{r, eval-definition, eval=FALSE}
base::eval <-
    function(expr, 
             envir = parent.frame(), 
             enclos = if (is.list(envir)||is.pairlist(envir)) parent.frame() else baseenv())
        .Internal(eval(expr, envir, enclos))

base::evalq <-
    function (expr, 
              envir = parent.frame(), 
              enclos = if (is.list(envir) || is.pairlist(envir)) parent.frame() else baseenv())
        .Internal(eval(substitute(expr), envir, enclos))

base::eval.parent <-
    function (expr, n = 1) {
        p <- parent.frame(n + 1)
        eval(expr, p)
    }

base::local <-
    function (expr, envir = new.env())
        eval.parent(substitute(eval(quote(expr), envir)))
```

```{r eval-count-table}
callsites_with_srcref <-
    eval_calls %>% 
    select(eval_call_srcref) %>% 
    n_distinct(na.rm = TRUE)

callsites_without_srcref <-
    eval_calls %>% 
    filter(is.na(eval_call_srcref)) %>% 
    select(caller_package, caller_function) %>% 
    n_distinct()

eval_calls <-
    eval_calls %>%
    mutate(eval_call_srcref = coalesce(eval_call_srcref, 
                                       str_c("missing", "::", caller_package, "::", caller_function)))
eval_count_table <-
    eval_calls %>%
    group_by(eval_function, eval_call_srcref) %>%
    summarize(count = n()) %>%
    summarize(call_count = sum(count), callsite_count = n())

nb_eval_calls <- total_call_count <- sum(eval_count_table$call_count)

nb_call_sites <- total_callsite_count <- sum(eval_count_table$callsite_count)

eval_count_table <-
    eval_count_table %>%
    mutate(call_proportion = 100 * call_count / total_call_count,
           callsite_proportion = 100 * callsite_count / total_callsite_count) %>%
    arrange(desc(call_count))

eval_call_proportion <- 
    eval_count_table %>%
    filter(eval_function == "eval") %>%
    pull(call_proportion)
```

There are `r NA` R function calls with `r total_call_count` calls to the four kinds of eval which accounts for `r NA`% of all R function calls.
The table below summarizes the proportion of calls to the different kinds of evals.

```{}
eval_count_table %>%
select(eval_function, call_count, call_proportion)
datatable()
```

We observe that `r eval_call_proportion`% of the calls are made to the `eval` function.

These `eval` calls originate from `r total_callsite_count` call sites. For the number of call sites, we use line number information attached to the AST by the parser. When it is not available (such as in *base*), we use a combination of `caller_package` and `caller_function`. We assume that:
  - the combination `caller_package` and `caller_function` is unique
  - there is only one call site per such `caller_function`
This assumption results in a lower bound for the number of unique callsites for eval.

```{r call-site-table}
eval_callsite_table <-
    eval_calls %>%
    count(eval_function, caller_package, eval_call_srcref, name = "call_count") %>%
    group_by(eval_function, caller_package) %>%
    summarize(call_count = sum(call_count), callsite_count = n()) %>%
    ungroup()

eval_callsite_table %>%
datatable()
```

```{r call-site-distribution}
package_callsite_distribution <-
    eval_callsite_table %>%
    mutate(callsite_count = if_else(callsite_count <= 5, as.character(callsite_count), "> 5")) %>%
    group_by(eval_function, callsite_count) %>%
    summarize(package_count = n())

package_callsite_distribution %>%
datatable()

p <-
    package_callsite_distribution %>%
    ggplot(aes(callsite_count, package_count)) +
    geom_col() +
    facet_wrap(vars(eval_function))
p
```

@@
